---
title: "Making age structure data by province"
output: html_document
date: "2023-02-24"
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Local folders

```{r}
gadm_data_folder <- "~/OneDrive - Oxford University Clinical Research Unit/data/GADM/"
wpop_data_folder <- "~/OneDrive - Oxford University Clinical Research Unit/data/WorldPop/raw/age structure/Vietnam"
wpop_data_processed <- "~/OneDrive - Oxford University Clinical Research Unit/data/WorldPop/processed/age structure/Vietnam/age_data.rds"
```


## Packages

Packages needed for the analysis:

```{r}
needed_package <- c("dplyr", "tidyr", "purrr", "terra", "stringr", "sf", "parallel", "magrittr")
```

Installing the packages that are not already installed:

```{r}
to_install <- needed_package[which(!needed_package %in% installed.packages()[, "Package"])]
if (length(to_install)) install.packages(to_install)
```

Loading the packages:

```{r}
dev_null <- sapply(needed_package, function(libname) library(libname, character.only = TRUE))
rm(dev_null)
```


## Downloading provinces polygons data from GADM

GADM root URL:

```{r}
gadm <- "https://geodata.ucdavis.edu/gadm/"
```

Version and format:

```{r}
vers_form <- "gadm3.6/Rsf/"
```

File:

```{r}
file_name <- "gadm36_VNM_1_sf.rds"
```

Creating the local folder structure if it does not exist:

```{r}
local_folder <- paste0(gadm_data_folder, vers_form)
if (!dir.exists(local_folder)) dir.create(local_folder, recursive = TRUE)
```

Downloading the data if not already available locally:

```{r}
local_file <- paste0(local_folder, file_name)
if (!file.exists(local_file)) download.file(paste0(gadm, vers_form, file_name), local_file)
```

Loading the polygons data:

```{r}
provinces <- local_file %>%
  readRDS() %>% 
  st_set_crs(4326) # setting the modern code of the WGS 84 projection
```

Checking the map of the provinces polygons:

```{r}
provinces %>% 
  st_geometry() %>% 
  plot()
```


## Downloading demographic raster data from WorldPop

WorldPop URL:

```{r}
worldpop <- "https://data.worldpop.org/"
```

All the combinations of years, ages and gender:

```{r}
ages <- c(0, 1, seq(5, 80, 5))

combinations <- expand_grid(years = 2000:2020,
                            ages  = ages,
                            sexes = c("f", "m"))
```

Creating the long files names:

```{r}
files_names <- with(combinations,
                    paste0("GIS/AgeSex_structures/Global_2000_2020/",
                           years, "/VNM/vnm_", sexes, "_", ages, "_", years, ".tif"))
```

The corresponding local files names:

```{r}
local_files_names <- paste0(wpop_data_folder, str_remove(files_names, "^GIS.*VNM"))
```

Checking for those that do not exist locally yet:

```{r}
sel <- which(!map_lgl(local_files_names, file.exists))
```

Downloading the files that are missing:

```{r}
if (length(sel)) {
  walk2(paste0(worldpop, files_names[sel]), local_files_names[sel], download.file)
}
```


## Generating province data

A function that converts the polygon of a given province of `provinces` into a
`SpatVector` object:

```{r}
prov2vect <- function(prov_name) {
  provinces %>% 
    filter(VARNAME_1 == prov_name) %>%
    vect()
}
```

A function that crops and masks a `SpatRaster` object `rst` with a `SpatVector`
object `pol`:

```{r}
crop_mask <- function(rst, pol) {
  rst %>%
    crop(pol) %>% 
    mask(pol)
}
```

A function that extracts the number of people of a raster `rst` inside a polygon
`pol`:

```{r}
nb_people <- function(rst, pol) {
  crop_mask(rst, pol) %>% 
    values() %>% 
    sum(na.rm = TRUE)
}
```

A function that converts a TIFF file (i.e. number of people for a given year,
age, and sex) into a data frame with number of people per province:

```{r}
tif2df <- function(tif_file) {
  tmp <- first(str_split(str_remove(tif_file, ".tif"), "_"))
  tif_file %>%
    paste0(wpop_data_folder, "/", .) %>% 
    rast() %>% 
    map_dbl(prov_vect_list, nb_people, rst = .) %>% 
    tibble(year = tmp[4], province = provinces_names, sex = tmp[2], age = tmp[3], n = .)
}
```

The names of the provinces:

```{r}
provinces_names <- provinces$VARNAME_1
```

A list of provinces polygons as `SpatVector`:

```{r}
prov_vect_list <- map(provinces_names, prov2vect)
```

Processing all the files to generate the age data frame (it takes about 30' on
11 cores of a 3.2 GHz 6-Core Intel i7):

```{r eval = FALSE}
hash <- setNames(c(paste0("[", head(ages, -1), ";", tail(ages, -1), "["), "80+"), ages)

age_data <- wpop_data_folder %>%
  dir() %>% 
  mclapply(tif2df, mc.cores = detectCores() - 1) %>% 
  bind_rows() %>% 
  mutate(age = hash[age],
         sex = c(f = "female", m = "male")[sex])
```

```{r include = FALSE}
age_data <- readRDS()
```


## Testing the data

Let's see at the change in the population size over the years:

```{r}
pop_in_a_year <- function(x) {
  age_data %>% 
    filter(year == x) %>% 
    pull(n) %>% 
    sum()
}

plot2 <- function(...) {
  plot(..., xlab = "", type = "o", col = 4)
}

years <- 2000:2020
pop_sizes <- map_dbl(years, pop_in_a_year)
plot2(years, pop_sizes, ylab = "population size")
```

Looks OK. Now, let's see proportion of different age classes:

```{r}
tmp <- age_data %>% 
  group_by(year, age) %>% 
  summarise(n = sum(n)) %>% 
  mutate(p = n / sum(n)) %>% 
  ungroup()

plot_age_prop <- function(x) {
  tmp %>% 
    filter(age == x) %$% 
    plot2(year, p, ylab = "proportion")
  mtext(x)
}

opar <- par(mfrow = c(6, 3))

age_data %>% 
  pull(age) %>% 
  unique() %>% 
  walk(plot_age_prop)

par(opar)
```

Looks OK too, in the sense that the proportions of the age classes are not
constant with time. Saving to disk:

```{r eval = FALSE}
saveRDS(age_data, "~/Desktop/age_data.rds")
```




