---
title: "Making age structure data by province"
output: html_document
date: "2023-02-24"
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(terra)
library(stringr)
library(raster)
library(sf)
library(parallel)
library(magrittr)
```

## Downloading provinces polygons data from GADM

GADM root URL:

```{r}
gadm <- "https://geodata.ucdavis.edu/gadm/"
```

Version and format:

```{r}
vers_form <- "gadm3.6/Rsf/"
```

File:

```{r}
file_name <- "gadm36_VNM_1_sf.rds"
```

Creating the local folder structure if it does not exist:

```{r}
local_folder <- paste0("~/OneDrive - Oxford University Clinical Research Unit/data/GADM/", vers_form)
if (!dir.exists(local_folder)) dir.create(local_folder, recursive = TRUE)
```

Downloading the data if not already available locally:

```{r}
local_file <- paste0(local_folder, file_name)
if (!dir.exists(local_file)) download.file(paste0(gadm, vers_form, file_name), local_file)
```

Loading the polygons data:

```{r}
provinces <- local_file %>%
  readRDS() %>% 
  st_set_crs(4326) # modern name of the WGS 84 projection
```

Checking the map of the provinces polygons:

```{r}
provinces %>% 
  st_geometry() %>% 
  plot()
```

## Downloading demographic raster data from WorldPop

WorldPop URL:

```{r}
worldpop <- "https://data.worldpop.org/"
```

Local data folder:

```{r}
data_folder <- "~/OneDrive - Oxford University Clinical Research Unit/data/WorldPop/"
```

```{r}
data_folder2 <- "~/OneDrive - Oxford University Clinical Research Unit/data/GADM 2/"
```

All the combinations of years, ages and gender:

```{r}
ages <- c(0, 1, seq(5, 80, 5))

combinations <- expand_grid(years = 2000:2020,
                            ages  = ages,
                            sexes = c("f", "m"))
```

Creating the files names:

```{r}
files_names <- with(combinations,
                    paste0("GIS/AgeSex_structures/Global_2000_2020/",
                           years, "/VNM/vnm_", sexes, "_", ages, "_", years, ".tif"))
```

Creating the folder structure within the data folder if does not exist:

```{r}
dir.create2 <- function(path, ...) {
  if (!dir.exists(path)) dir.create(path, recursive = TRUE, ...)
}

files_names %>% 
  map_chr(str_remove, "/v.*$") %>% 
  unique() %>% 
  paste0(data_folder, .) %>% 
  walk(dir.create2)
```

The remote and local files paths:

```{r}
urls <- paste0(worldpop, files_names)
destfiles <- paste0(data_folder, files_names)
```

Downloading the data:

```{r}
download.file2 <- function(url, destfile, ...) {
  if (!file.exists(destfile)) download.file(url, destfile, ...)
}

walk2(urls, destfiles, download.file2)
```

## Generating province data

A function that converts the polygon of a given province of `provinces` into a
`SpatVector` object:

```{r}
prov2vect <- function(prov_name) {
  provinces %>% 
    filter(VARNAME_1 == prov_name) %>%
    vect()
}
```

A function that crops and masks a `SpatRaster` object `rst` with a `SpatVector`
object `pol`:

```{r}
crop_mask <- function(rst, pol) {
  rst %>%
    crop(pol) %>% 
    mask(pol)
}
```

A function that extracts the number of people of a raster `rst` inside a polygon
`pol`:

```{r}
nb_people <- function(rst, pol) {
  crop_mask(rst, pol) %>% 
    values() %>% 
    sum(na.rm = TRUE)
}
```

A function that converts a tif file (i.e. number of people for a given year,
age, and sex) into a data frame with number of people per province:

```{r}
tif2df <- function(tif_file) {
  tmp <- first(str_split(str_remove(tif_file, ".tif"), "_"))
  tif_file %>%
    paste0(data_folder2, .) %>% 
    rast() %>% 
    map_dbl(prov_vect_list, nb_people, rst = .) %>% 
    tibble(year = tmp[4], province = provinces_names, sex = tmp[2], age = tmp[3], n = .)
}
```

The names of the provinces:

```{r}
provinces_names <- provinces$VARNAME_1
```

A list of provinces polygons as `SpatVector`:

```{r}
prov_vect_list <- map(provinces_names, prov2vect)
```

Processing all the files to generate the age data frame:

```{r}
age_data <- data_folder2 %>%
  dir() %>% 
  head(3) %>% 
  map(tif2df) %>% 
  bind_rows()
```

A parallelized version:

```{r}
hash <- setNames(c(paste0("[", head(ages, -1), "; ", tail(ages, -1), "["), "80+"), ages)

age_data <- data_folder2 %>%
  dir() %>% 
#  head(3) %>% 
  mclapply(tif2df, mc.cores = detectCores() - 1) %>% 
  bind_rows() %>% 
  mutate(age = hash[age]) %>% 
  mutate(sex = c(f = "female", m = "male")[sex])
```

## Testing

Let's see at the change in the population size over the years:

```{r}
pop_in_a_year <- function(x) {
  age_data %>% 
    filter(year == x) %>% 
    pull(n) %>% 
    sum()
}

years <- 2000:2020
pop_sizes <- map_dbl(years, pop_in_a_year)
plot(years, pop_sizes, type = "o", xlab = "", ylab = "population size")
```

Works good! Now, let's see proportion of different age classes:

```{r}
tmp <- age_data %>% 
  group_by(year, age) %>% 
  summarise(n = sum(n)) %>% 
  mutate(p = n / sum(n)) %>% 
  ungroup()

plot_age_prop <- function(x) {
  tmp %>% 
    filter(age == x) %$% 
    plot(year, p, type = "o")
}

age_data %>% 
  pull(age) %>% 
  unique() %>% 
  walk(plot_age_prop)
```

```{r}
saveRDS(age_data, "age_data.rds")
```




